StarCraft Bots and Competitions

### 2015-04-22 ECGG_StarCraftCompetitions/ECGG_StarCraftCompetitions.tex
Init SC Competitions paper

\usepackage{algorithm}
\usepackage{fixltx2e}
\usepackage{algpseudocode}

@@

\section{Definition}

{\color{blue} to-do}

>>>>

\usepackage{algorithm}
\usepackage{fixltx2e}
\usepackage{algpseudocode}
\usepackage{adjustbox}

@@

\section{State of the Art Bots for StarCraft}\label{sec:bot}

Thanks to the recent organization of international game AI competitions focused around the popular StarCraft game (see Section \ref{sec:competition}), several groups have been working on integrating many of the techniques described in the previous section into complete ``bots'', capable of playing complete StarCraft games. In this section we will overview some of the currently available top bots.

%\subsection{RTS Bot Architectures}\label{sec:integration}

Playing an RTS game involves dealing with all the problems described above. A few approaches, like CAT \cite{LTW}, Darmok \cite{OntanonMSR10} or ALisp \cite{Marthi05} try to deal with the problem in a monolithic manner, by using a single AI technique. However, none of those systems aims at achieving near human performance. In order to achieve human-level performance, RTS AI designers use a lot of domain knowledge in order to divide the task of playing the game into a collection of sub-problems, which can be dealt-with using individual AI techniques.

\begin{figure}[h!]
  \begin{adjustbox}{addcode={\begin{minipage}{\width}}{\caption{%
    %\centering
            Architecture  of 7  StarCraft bots  obtained by  analyzing
            their  source code.  Modules  with  black background  sent
            commands  directly to  StarCraft, dashed  arrows represent
            data flow, and solid arrows represent control.
          }\end{minipage}},rotate=90,center}
    \includegraphics[width=0.935\textheight]{figures/figure-bot-architectures-wide.pdf}
    % \caption{Architecture of 7  StarCraft bots obtained by analyzing
    % their  source code.  Modules with  black background  sent commands
    % directly  to StarCraft,  dashed  arrows represent  data flow,  and
    % solid arrows represent control.}
  \end{adjustbox}
    \label{fig:bot-architecture}
\end{figure}

Figure \ref{fig:bot-architecture} shows some representative examples of the architectures used by different bots in the AIIDE and CIG StarCraft AI competitions (see Section \ref{sec:competition}): BroodwarBotQ \cite{SynnaeveMicroCig11}, Nova \cite{uriarte2012kiting}, UAlbertaBot \cite{churchill2011build}, Skynet, SPAR, AIUR, and BTHAI \cite{Hagelback12}. Each box represents an individual module with a clearly defined task (only modules with a black background can send actions directly to StarCraft). Dashed arrows represent data flow, and solid arrows represent control (when a module can command another module to perform some task). For example, we can see how SPAR is divided in two sets of modules: {\em intelligence} and {\em decision making}. Intelligence in SPAR has three modules dedicated to analyze the current situation of the game. Decision making in SPAR is done through four hierarchically organized modules, with the higher-level module ({\em strategic decision}) issuing commands to the next module ({\em tactical decision}), which sends commands to the next module ({\em action implementation}), and so on. Only the two lower-level modules can send actions directly to StarCraft.

On the other hand, bots such as Nova or BroodwarBotQ (BBQ) only use a hierarchical organization for {\em combat} (controlling the attack units), but use a decentralized organization for the rest of the bot. In Nova and BBQ, there is a collection of modules that control different aspects of the game (workers, production, construction, etc.). These modules can all send actions directly to StarCraft. In Nova those modules coordinate mostly through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator: a bidding market and broker for settling units control, military and civilian groups/task forces bid for units proportionally to their usefulness and the task importance.

By analyzing the structure of these bots, we can see that there are two main tools being used in these integration architectures:

\begin{itemize}
\item {\em Abstraction}: complex tasks can be formulated at different levels of abstraction. For example, playing an RTS game can be seen as issuing individual low-level actions to each of the units in the game, or at a higher level, it can be seen as deploying a specific strategy (e.g. a ``BBS strategy'', or a ``Reaver Drop'' strategy). Some bots, reason at multiple levels of abstraction at the same time, making the task of playing StarCraft simpler. Assuming that each module in the architecture of a bot has a goal and determines some actions to achieve that goal, the actions determined by higher-level modules are considered as the goals of the lower level modules. In this way, each module can focus on reasoning at only one level of abstraction, thus, making the problem easier.

\item {\em Divide-and-conquer}: playing a complex RTS, such as StarCraft, requires performing many conceptually different tasks, such as gathering resources, attacking, placing buildings, etc. Assuming each of these tasks can be performed relatively independently and without interference, we can have one module focusing on each of the tasks independently, thus making the problem easier. 
\end{itemize}

If we imagine the different tasks to perform in a complex RTS game in a two-dimensional plane, where the vertical axis represents abstraction, and the horizontal axis represents the different aspects of the game (combat, resource gathering, etc.), abstraction can be seen as dividing the space with horizontal lines, whereas divide-and-conquer divides the space using vertical lines.

Different bots use different combinations of these two tools. Looking back at Figure \ref{fig:bot-architecture}, we can see the following use of abstraction and divide-in-conquer in the bots:

\begin{itemize}
\item
  BroodwarBotQ\footnote{\url{http://github.com/SnippyHolloW/BroodwarBotQ}}:
  uses  abstraction for  {\em combat}, and  divide-and-conquer for
  {\em economy}  and  {\em intelligence  gathering}. To  avoid  conflicts
  between modules (since  the individual tasks of each  of the modules
  are not completely independent), BBQ uses an arbitrator.
\item  Nova\footnote{\url{http://nova.wolfwork.com/}}:  is similar  in
  design as  BroodwarBotQ, and uses  abstraction for {\em combat},
  and  divide-and-conquer for  {\em economy}.  The differences  are
  that Nova does not have  an arbitrator to resolve conflicts, but has
  a  higher-level   module  ({\em  strategy   manager}),  which  posts
  information to the blackboard that the rest of modules follow (thus,
  making use of abstraction).
\item
  UAlbertaBot\footnote{\url{http://code.google.com/p/ualbertabot/}}:
  also  uses abstraction  in  {\em combat} like  the previous  two
  bots. But it  also uses it in {\em economy}: as  can be seen, the
  production manager sends commands to the building manager, who is in
  charge   of   producing   the   buildings.  This   bot   also   uses
  divide-and-conquer, and  tasks like scouting  and resource gathering
  are managed by separate, independent modules.
\item       Skynet\footnote{\url{http://code.google.com/p/skynetbot/}}:
  makes     extensive     use      of     both     abstraction     and
  divide-and-conquer.  We  can see a  high level module  that issues commands  to a
  series of  tactics modules. The  collection of tactic  modules queue
  {\em  tasks} (that  are analogous  to the  abstract actions  used in
  SPAR).  Each different  task has  a specific  low level  module that
  knows how to  execute it. Thus, Skynet uses  a 3 layered abstraction
  hierarchy,  and uses  divide-and-conquer  in all  levels except  the
  highest.
\item
  SPAR\footnote{\url{http://www.planiart.usherbrooke.ca/projects/spar/}}:
  only uses abstraction. Its high-level module determines the strategy
  to  use,  and  the  tactical  decision  module  divides  it  into  a
  collection  of {\em  abstract  actions}, that  are  executed by  the
  lower-level modules.
\item  AIUR\footnote{\url{http://code.google.com/p/aiurproject/}}:  is
  mainly divide-and-conquer oriented,
  with a  slight abstraction on  {\em economy} due to a  SpendManager deciding
  how  to  spend  and  share  resources  among  Base,  Production  and
  Construction Managers.  At the beginning of a  game, the MoodManager
  initializes  a  ``mood''  which  will  influence  both  tactics  and
  strategy. {\em Combat} is divided into three  independent managers: the
  {\em  Defense Manager},  controlling  military units  when there  is
  nothing special, the {\em  Under Attack Manager}, activated when the
  opponent is attacking our bases,  and the {\em Army Manager}, taking
  control of units when it is time to attack, following a timing given
  by the  current mood. This bot  does not manage however  any kind of
  reactive controls so far.
\item  BTHAI\footnote{\url{http://code.google.com/p/bthai/}}:  uses  a
  two-tier  abstraction hierarchy,  where a  collection  of high-level
  modules command a collection of lower-level agents in charge of each
  of  the units.  At  the high-level,  BTHAI uses  divide-and-conquer,
  having   multiple  high-level  modules   issuing  commands   to  the
  lower-level units.
\end{itemize}

Additionally, except for BTHAI, all other agents use divide-and-conquer at a higher-level bot design and divide all the modules into two or three categories: {\em intelligence gathering} and {\em decision making} (sometimes divided into {\em combat} and {\em economy}).

Some bots  using divide-and-conquer, assume  that each of  the modules
can act independently  and that their actions can  be executed without
interference.  BBQ,  UAlbertaBot and  AIUR, however use  an arbitrator
({\em Game Commander'} in UAlbertaBot) that makes sure that modules do
not send contradictory orders to  the same unit.  However, very little
bots handle  the problem of  how to coordinate resource  usage amongst
modules, for  instance BTHAI uses a  first-come-first-serve policy for
spending resources,  the first module  that requests resources  is the
one that gets them. Nova and Skynet are exceptions, and implement some
rudimentary prioritization based on the high level strategy. Following
available  resources and  timing,  AIUR's {\em  Spend Manager}  orders
Base,  Production   and  Construction  Managers  what   they  have  to
build/produce. It also orders to start tech research and upgrades. The
idea here is not to let the different managers allocate the resources they want, but to do the opposite, that is: finding how the AI can
spend the available money.

One  interesting aspect  of the  seven bots  described above  is that,
while all  of them (except AIUR) are  reactive at the  lower level
(reactive  control), most  if not  all of  them, are  scripted  at the
highest level  of abstraction. BTHAI reads build  and squad formations
from  a  predefined  script,   Nova's  {\em  Strategy  Manager}  is  a
predefined finite-state machine,  BBQ's construction manager reads the
build  order from a  predefined script,  and Skynet's  {\em BuildOrder
  Manager} is basically a predefined script. Such scripts describe the
strategy  that the  bots will  use, however,  such strategy  is always
fixed.   One could see  this pre-scripting  as if  each bot  defined a
``high-level programming language''  to describe StarCraft strategies,
and   the   bots   themselves    are   just   interpreters   of   such
strategy.  Compared  to  current  approaches  for Chess  or  Go,  this
scripting  seems a  rigid and  inflexible,  but responds  to the  much
higher complexity of the  StarCraft game.  An interesting exception to
that  is  UAlbertaBot, which  uses  a  search  algorithm in  the  {\em
  Production  Manager}  to find  near-optimal  build orders.   Another
interesting case is  AIUR, that uses a {\em  Mood Manager} to randomly
pick  a mood among  six (cheese,  rush, aggressive,  defensive, macro,
fast  expand), which  will  influence the  build  order, strategy  and
tactics. % All of  them  scripted so  far,  but build  orders can  been modified but  the Spend Manager, following  available resources.  This mood can change during the game, regarding the opponent behavior.

In conclusion, we can see that there are two basic tools that can be used in an integration architecture: abstraction and divide-and-conquer, which are widely used by the existing StarCraft bots. For space reasons, we do not include an exhaustive comparison of the architectures of all the participating bots. Some other bots have been documented by their authors, such as SCAIL \cite{YoungSCAIL} or QUORUM \cite{young2012evolutionary}.
%This section has focused on discussing the architecture of existing StarCraft bots. 
Let us now focus on their performance.

%\subsection{Individual AI Techniques}\label{sec:techniques}

%{\color{red} TODO}

\section{Recent StarCraft AI Competitions}\label{sec:competition}

This section reviews the results of the recent international competitions on AI for StarCraft. These competitions, typically co-located with scientific conferences, have been possible thanks to the existence of the Brood War Application Programming Interface 
(BWAPI)\footnote{\url{http://code.google.com/p/bwapi/}}, which 
enables replacing the human player interface with C++ code.
%A consequence of this indirect way of integrating new bots is that every machine can only run one custom bot, requiring two computers for a 2-player game.
The following subsections summarize the results of all the StarCraft AI competitions held at the AIIDE (Artificial Intelligence for Interactive Digital Entertainment) and CIG (Computational Intelligence in Games) conferences during the past years. Additionally we analyze the statistics from the StarCraft Bot Ladder, where the best bots play against each other continuously over time.

\subsection{AIIDE}\label{sec:AIIDE}

Started in 2010, the AIIDE StarCraft AI Competition\footnote{\url{http://www.StarCraftAICompetition.com}} is the most
well known and longest running StarCraft AI Competition in the world. Each year,
AI bots are submitted by competitors to do battle within the retail version of
StarCraft: Brood War, with prizes supplied by Blizzard Entertainment. %This competition has been made possible by the BWAPI StarCraft programming interface, which allows players to write programs which can retrieve game data and issue game commands to StarCraft via an easy to use C++ API.

The first competition in 2010 was organized and run by Ben Weber in the Expressive
Intelligence Studio at University of California, Santa Cruz\footnote{\url{http://eis.ucsc.edu/StarCraftAICompetition}}. 26 total submissions
were received from around the world. As this was the first year of the competition,
and little infrastructure had been created, each game of the tournament was run 
manually on two laptop computers and monitored by hand to record the results. Also,
no persistent data was kept for bots to learn about opponents between matches.

The 2010 competition had 4 different tournament categories in which to compete. Tournament 1
was a flat-terrain unit micro-management battle consisting of four separate unit
composition games. Of the six competitors, FreSCBot won the competition with
Sherbrooke coming in 2nd place. Tournament 2 was another micro-focused game with
non-trivial terrain. Two competitors submitted for this category, with FreSCBot
once again coming in 1st by beating Sherbrooke.

Tournament 3 was a tech-limited StarCraft game on a single known map with no fog-of-war enforced. Players
were only allowed to choose the Protoss race, with no late game units allowed. 8 bots
faced off in this double-elimination tournament with MimicBot taking first place over Botnik in the final.
As this was a perfect information variant of StarCraft, MimicBot adopted a strategy of
``mimic its opponent's build order, gaining an economic advantage whenever possible'' which
worked quite well.

Tournament 4 was the complete game of {\em StarCraft: Brood War} with fog-of-war enforced. The tournament
was run with a random pairing double-elimination format with each match being best of 5 games.
Competitors could play as any of the three races, with the only limitations in gameplay
being those that were considered ``cheating'' in the StarCraft community. A map pool of
5 well-known professional maps were announced to competitors in advance, with a random map being chosen for each game.

Results are shown in Table \ref{tab:aiide2010}. The team that won was Overmind\footnote{\url{http://overmind.cs.berkeley.edu}}, from University of California, Berkeley. Using the Zerg race, their
strategy was to defend early aggression with zergling units while amassing mutalisk units,
which they used to contain and eventually defeat their opponents. The mutalisk is a very fast and agile
flying unit which is able to attack while moving with no drawback, which makes them quite a powerful
unit when controlled by a computer. Overmind used a potential-field
based micro-management system to guide their mutalisks, which led them to victory. Krasi0 came in 2nd place
with a standard defensive Terran opening strategy that transitioned into ``mech'' play in the late game.

\begin{table}[!t]
\caption{Ranking of the three best bots of the AIIDE 2010 competition}
\label{tab:aiide2010}
\centering
\begin{tabular}{|c|l|}
\hline
{\bfseries Position} & {\bfseries Bot}\\
\hline
1 & Overmind \\
2 & Krasi0 \\
3 & Chronos \\
\hline
\end{tabular}
\end{table}

In 2011 the University of Alberta hosted the competition, with organization by Michael Buro and
David Churchill\footnote{\url{https://skatgame.net/mburo/sc2011/}}. Due to a lack of entrants in tournament categories 1-3 in the 2010 competition, it was
decided that only the full game category would be played in the 2011 competition. Another important
change in the 2011 competition was the introduction of automated tournament-managing software running
StarCraft games simultaneously on 20 computers, allowing a total of 1170 games to be played in far less
time than the 108 games of the 2010 competition. This increase in games played also allowed the tournament
to switch to a round-robin format, eliminating the ``luck'' factor of the pairings inherent in bracket
style tournaments. The bot that achieved the highest win percentage over the course of the competition would
be determined the winner. Also, the competition became open-source, in an effort not
only to prevent possible cheating, but to promote healthy competition in future tournaments by giving
newcomers and easier entry point by basing their design off of previous bots.

In the  end, Skynet  won the competition  with its solid  Protoss play
(results  are  summarized   in  Table  \ref{tab:aiide2011}).  The  bot
executed one of a small set of strategies randomly at the start of the
match based on the map and the race of the opponent. Skynet would then
amass a  medium to large  sized army and  expand before moving  out to
attack. Good  use of Dragoon (powerful ranged ground unit with clumsy movement) range and  kiting micro-management allowed
it to hold off the early aggression of other bots such as UAlbertaBot,
which came in 2nd.

\begin{table}[!b]
\caption{Results of the five best bots of the AIIDE 2011 competition}
\label{tab:aiide2011}
\centering
\begin{tabular}{|c|l|r|}
\hline
{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
\hline
1 & Skynet & 88.9\% \\
2 & UAlbertaBot & 79.4\% \\
3 & AIUR & 70.3\% \\
4 & ItayUndermind & 65.8\% \\
5 & EISBot & 60.6\% \\
\hline
\end{tabular}
\end{table}

UAlbertaBot used an early zealot-rush strategy to take advantage of the power of early game Protoss units. 
It would send out the first zealots that were made and immediately attack the enemy base, using a unit
counting heuristic to determine whether or retreat or keep pushing. Of note is that UAlbertaBot used an
online planning algorithm to construct all of its economic build-orders\cite{churchill2011build}, as no hard-coded build orders
were used.

AIUR also chose  Protoss, with a strategy that was in between Skynet
and UAlbertaBot in terms of  attack timings.  At that time, AIUR chose
one  mood among  five (leading  to slightly  different  strategies and
tactics) at the  beginning of a game and kept it  until the end. These
five moods were:
\begin{itemize}
\item  Rush:  where  the  bot  tries  early  attacks,  and  have  good
  probabilities to send the two or three first Zealots (basic contact attack ground unit) to harass the
  opponent.
\item Aggressive: where  we have less chance to  perform harasses with
  the first Zealots, and the first attack is usually a bit delayed
  with regard to the Rush mood.
\item Macro: where the AI do not try any early attacks and focus a bit
  more on its economy before attacking.
\item Defense: where the AI  ``turtles'' and wait to have a consequent
  army before running an attack.
\item  Fast expand:  where the  first building  constructed it  a base
  expansion, for a very economical-oriented game.
\end{itemize}
Notice that  build orders are not  fully hard-coded since  they can be
altered by AIUR's Spend Manager.

Of  note  in  these  results  was that  a  rock-paper-scissors  effect
happened among  the top  3 finishers.  Of  the 30 rounds,  Skynet beat
UAlbertaBot 26  times, UAlbertaBot beat  AIUR 29 times, and  AIUR beat
Skynet  19 times.   Another notable  result is  that Overmind  did not
choose  to compete despite  winning the  2010 competition.   After the
competition,  many  bot  programmers  (including  the  Overmind  team)
realized that their  2010 strategy was quite easily  defeated by early
game rushing strategies,  and so they submitted a  Terran bot instead,
called Undermind, which finished in 7th.

After the competition was over, a man vs. machine match was held between the winner (Skynet) and an
ex-professional StarCraft player named Oriol Vinyals. Oriol was a competitor in the 2001 World Cyber
Games StarCraft competition, and though he had been out of practice for a few years was still quite a
good player. The match was arranged to see how well StarCraft AI bots had progressed and to see if
they could actually beat a decent human opponent.

\begin{figure}[t!]
    \centering
    \includegraphics[width=\columnwidth]{figures/aiide2012_v2}
    \caption{Evolution of the win percentage of each bot participating in the AIIDE 2012 competition}
    \label{fig:aiide2012}
\end{figure}

For the best-of-three match, Oriol chose his races randomly and ended up beating Skynet in a 2-0. 
In the first match, Oriol played Zerg vs. Skynet's Protoss on Python, a four player map. Oriol chose
to start with a fast expansion strategy and transition into two base mutalisk production. Skynet chose
to rush with a few early zealots, which was luckily the best possible choice given Oriol's strategy.
Skynet's initial attack destroyed Oriol's early defenses, and nearly won the game in the first few
minutes, however it then proceeded to send zealots to attack one at a time rather than group up its
units before moving in, which allowed Oriol to catch up. Once Oriol produced his mutalisks, Skynet did
not produce sufficient air defenses and Oriol quickly destroyed Skynet's base. In the second game,
Oril played Terran, again on Python. After holding off early Dragoon pressure from Skynet, Oriol
moved out with a few marines, medics and tanks. Skynet tried to defend with its army of Dragoons,
however due to poor unit targeting decisions it started to attack useless medics after the marines
had died, rather than the tanks. Oriol overcame the Dragoon army and was victorious. Later analysis
of the match concluded that Skynet, while dominant over the other bots, was unable to properly
adapt and transition into a mid-game strategy in game one once its early pressure failed, and in game
two made a key blunder in unit targeting which cost it the game. Humans were still in command.

\begin{table}[!t]
\caption{Results of the five best bots of the AIIDE 2012 competition}
\label{tab:aiide2012}
\centering
\begin{tabular}{|c|l|r|}
\hline
{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
\hline
1 & Skynet & 84.4\% \\
2 & AIUR & 72.2\% \\
3 & UAlbertaBot & 68.6\% \\
4 & BroodwarBotQ & 59.1\% \\
5 & AdjutantBot & 52.8\% \\
\hline
\end{tabular}
\end{table}

The University of Alberta also hosted the 2012 competition, with the major difference from the 2011
competition being the addition of persistent storage. Bots could now write information to disk during a
match, and then read the information during other matches, allowing them to adjust strategies based
on previous results. 6 of the 10 entrants used this feature to aid in strategy selection, including the
top 4 finishers. More improvements to the tournament environment also meant that a total of 4240 games
could now be played in the same time period. Results are shown in Table \ref{tab:aiide2012}.

Skynet once  again won  the competition with  its solid  Protoss build
orders  and  good  Dragoon  kiting.   AIUR  and  UAlbertaBot  switched
positions from  the previous  year to come  2nd and  3rd respectively.
Both  AIUR  and UAlbertaBot  used  data  stored  from the  results  of
previous games  to select a strategy for  future matches.  UAlbertaBot
did  this  using  the  UCB  \cite{auer2002finite} algorithm,  while  AIUR  used  a  uniform
distribution  to choose  its  mood before  altering this  distribution
after  some  games  against  the  same  opponent  to  favor  efficient
strategies, achieving  similar results than  UAlbertaBot. Notice that,
compared  to AIIDE 2011,  AIUR proposes  a new  mood, \textit{Cheese},
implementing a  Photon Cannon rush  strategy in order to  surprise the
opponent and  to finish the game  as soon as possible.   The effect of
this strategy selection process can be seen Figure \ref{fig:aiide2012}
which shows bot  win percentages over time.  While  the earlier rounds
of the tournament fluctuated wildly in results, eventually the results
converged to their  final values. One of the main  reasons for this is
due to  the bots  learning which strategies  to use as  the tournament
progressed.

The 2012 man vs. machine match again used the winner of the competition (Skynet), who played against
Mike Lange, also known as Bakuryu. At the time of the match, Bakuryu was an A- ranked Zerg player on ICCup,
and known as one of the best non-Korean Zerg players in the world. Bakuryu was considered much
stronger than Oriol at the time that the match was played, and the results showed that this was true.
In the first game of the best-of-three, Bakuryu made Skynet look quite silly by running around inside
Skynet's base with a small number of zerglings while Skynet's zealots and half of its worked chased
then in vain. After killing off several probes and buying enough time to set up his expansion, he
cleaned up Skynet's army with a flying army of mutalisks. In the second game, Bakuryu contained Skynet
inside its base with a group of zerglings positioned within Skynet's expansion. Skynet then constructed
several Dark Templar and along with some Dragoons and Zealots attacked into Bakuryu's expansion which
was heavily defended, and was crushed almost instantly, allowing Bakuryu's zergling force to finish
off the Protoss base.

In this match it was shown that the true weakness of state of the art StarCraft AI systems was that
humans are very adept at recognizing scripted behaviors and exploiting them to the fullest. A human
player in Skynet's position in the first game would have realized he was being taken advantage of and
adapted his strategy accordingly, however the inability to put the local context (Bakuryu kiting his
units around his base) into the larger context of the game (that this would delay Skynet until
reinforcements arrived) and then the lack of strategy change to fix the situation led to an easy
victory for the human. These problems remain as some of the main challenges in RTS AI today: to both recognize
the strategy and intent of an opponent's actions, and how to effectively adapt your own strategy to 
overcome them.

All results, videos, and replays from the AIIDE StarCraft AI Competition can be found in \url{http://www.StarCraftAICompetition.com}.
% \footnote{\url{http://www.StarCraftAICompetition.com}}.

\begin{table*}[!t]
\caption{Results of the first round at CIG 2011, held in two brackets.
Qualified for the final round: UAlbertaBot and Skynet (from A), Xelnaga
and BroodwarBotQ (from B, the latter by comparing direct encounters
with BTHAI of which 6:4 were won)}
\label{tab:cig-first-round}
\centering
% bracket A
\begin{tabular}{|c|c|c|l|r|}
\hline
\multicolumn{5}{|c|}{Bracket A} \\ \hline
{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
\hline
A1 & 0 &   40 &  UAlbertaBot & 82.5\% \\
A2 & 1 &     40 &  Skynet   &  77.5\% \\
A3 & 2 &   40 &  AIUR   &  60.0\% \\
A4 & 1 &   40 &  Nova   &  20.0\% \\
A5 & 0 &   40 &  LSAI   &  10.0\%\\
\hline
\end{tabular}
% bracket B
\begin{tabular}{|c|c|c|l|r|}
\hline
\multicolumn{5}{|c|}{Bracket B} \\ \hline
{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
\hline
B1 & 12 &  40 &  Xelnaga &   62.5\%\\
B2 & 3 &   40 &  BroodwarBotQ  &  57.5\%\\
B3 & 0 &   40 &  BTHAI    &  57.5\%\\
B4 & 17 &  40 &  Protoss Beast Jelly  & 42.5\%\\
B5 & 0 &   40 &  EvoBot   &  30.0\%\\
\hline
\end{tabular}
\end{table*}

\subsection{CIG}
\label{sec:cig2011}

An initial attempt to run a StarCraft tournament at the Computational
Intelligence in Games conference (CIG 2010) suffered from technical problems.
These mainly stemmed from the desire to use evolved, largely untested
maps which proved to look interesting but made the submitted bots and the Brood War Terrain Analyzer (BWTA) provided with the BWAPI interface crash
so frequently that it would have been unjustifiable to announce a winner.

At CIG 2011, the tournament was therefore run with a (secret) selection
of maps used in league play, which can be regarded as the most important
difference to the AIIDE tournament that employed a known list of maps.
The competition was organized by Tobias Mahlmann and Mike
Preuss and attracted 10 bots. In addition to the ones discussed in previous
sections (UAlbertaBot, Skynet, AIUR, Nova, BroodwarBotQ, BTHAI), 
the set also contained LSAI, Xelnaga, Protoss Beast Jelly, and EvoBot,
these are shortly described in the following:

\paragraph*{LSAI (Zerg)} utilizes a heavily modified BWSAL\footnote{\url{https://code.google.com/p/bwsal/}} to divide management
 of the units to different modules that communicate via a centralized information
 module. It works using a simple reactive strategy to try and survive early game
 attacks and macro up to a larger attack force and maintain map control.

\paragraph*{Xelnaga (Protoss)} is a modification of the AIUR bot that chooses the 
Dark Templar Opening in order to destroy the enemy base before defenses against
invisible units are available.

\paragraph*{Protoss Beast Jelly (Protoss)}
always goes for a 5-gate Zealot rush, supported by an effective harvesting 
strategy named power-mining (2 probes are assigned to every mineral patch,
thereby needing 18 probes for 100\% saturation in a normal map, prior
to expanding). Gas is not mined as it is not needed for constructing Zealots.

\paragraph*{EvoBot (Terran)} employs an evolutionary algorithm for obtaining rational
 unit combinations and influence map techniques for deciding the strategic locations. Note
 that this bot was submitted in a very early version, with many of its designed features not
 yet fully ready.

\subsubsection{First Round}
\label{sec:cig-first-round}

\begin{figure}[t]
	%\vspace*{-1ex}
    \centering
    \includegraphics[width=\columnwidth]{figures/cig2012-ResultsRound90.pdf}
	%\vspace*{-1ex}
    \caption{Evolution of the win percentage of each bot participating in the
CIG 2012 competition}
    \label{fig:cig2012-results}
\end{figure}

\begin{figure*}[b]
	%\vspace*{-2ex}
    \centering
    \includegraphics[width=0.32\textwidth]{figures/vstable3.pdf}
    \includegraphics[width=0.32\textwidth]{figures/vstable6.pdf}
    \includegraphics[width=0.32\textwidth]{figures/vstable.pdf}
	%\vspace*{-4ex}
    \caption{Win percentages of CIG 2012 competition, from left
    to right: 3-player maps only, 6-player maps only, all maps.
    Read from line to column, bot in row wins given fraction of
    games against bot in column. For some bots, we find interesting
    differences, e.g. Xelnaga gets worse on 6-player maps, UAlbertaBot
    gets better. Only Xelnaga can reliably beat Skynet, but only on 
    3-player maps}
    \label{fig:cig2012-mapresults}
\end{figure*}

As the CIG competition games were executed manually due to
a lack of available software (the AIIDE program was not yet
available at that time), the organizers separated the ten entries into
two brackets. In each bracket of 5 bots, a round-robin
tournament was held with 10 repetitions per pairing, resulting
in 40 games per bot.
The 5 maps chosen for the first round were selected from the pool
of well-known league play maps found on the Internet:
\emph{(2)MatchPoint 1.3}, \emph{(4)Fighting Spirit 1.3}, 
\emph{iCCup Destination 1.1}, \emph{iCCup Gaia}, and 
\emph{iCCup Great Barrier Reef}. Each bot pairing played
on every map twice, with switched starting positions.

The two top bots of every bracket qualified
for the final round. Table~\ref{tab:cig-first-round} summarizes
the results.
Note that as BroodwarBotQ and BTHAI have the same number of wins,
their direct encounter was evaluated which accounted 6:4 for the BroodwarBotQ.
The bots going into the final were thus UAlbertaBot, Skynet (from bracket A)
and Xelnaga and BroodwarBotQ (from bracket B). All qualified bots play the
Protoss faction. Most bots proved pretty stable, only Xelnaga and Protoss 
Beast Jelly crashed relatively often (each in more than a quarter of the games). 
Crashing of course resulted in an instant win for the other bot.
In some cases, neither bot was able to finish the other off completely,
so that they went into a passive state. We manually ended such games after
around 15 minutes and assigned victory to the bot that had obtained more
points as indicated on the end game screen.

\begin{table}[!b]
\caption{Results of the CIG 2011 competition}
\label{tab:cig-final-round}
\centering
\begin{tabular}{|c|c|c|l|r|}
\hline
{\bfseries Position} & {\bfseries Crashes} & {\bfseries Games} & {\bfseries Bot} & {\bfseries Win \%} \\
\hline
1 & 0 &  30 &  Skynet       &  86.7\%\\
2 & 0 &  30 &  UAlbertaBot  &  73.3\%\\
3 & 3 &  30 &  Xelnaga      &  36.7\%\\
4 & 2 &  30 &  BroodwarBotQ   &  3.3\%\\
\hline
\end{tabular}
\end{table}

\subsubsection{Final Round}
\label{sec:cig-final-round}

The final round was played in a similar mode as each of the
first round brackets,
using another set of 5 previously unknown maps:
\emph{iCCup lost temple 2.4}, \emph{iCCup rush hour 3.1},
\emph{iCCup swordinthemoon 2.1}, \emph{iCCup yellow 1.1},
and \emph{La\_Mancha 1.1}. 
Letting each pairing play on each map twice again with
switching starting positions resulted in 30 games per bot.
The final results are displayed in table~\ref{tab:cig-final-round},
indicating Skynet as winner and UAlbertaBot as runner-up, being
almost equally strong, and the two other bots as clearly inferior.
The competition setup, documentation and results can be found
in\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/StarCraft-cig2011}}.

For CIG 2012, the AIIDE tournament software was employed,
leading to a total of $4050$ games played in $90$ rounds
of round robin. As 6 different maps were used, this means
that each bot played every other on every map 15 times. 
As in the AIIDE competition, writing to and reading from
a bot specific directory was enabled, however, due to 
technical reasons, this feature was constrained to the
computer (of 6) the game was actually run on. We can 
therefore assume that this feature was of minor use for
the CIG competition. The only other difference to the
AIIDE competition was that the used maps were not 
made available to the competitors in advance.

These maps came in two flavors, namely three 3-player maps:
\emph{Athena-II}, \emph{Neo Moon Glaive}, \emph{Tears of the Moon},
and three 6-player maps:
\emph{Legacy}, \emph{River of Light}, and 
\emph{The Huntress 1.1}. 
We shall note that some bots consistently crashed on one
of the originally considered maps which has thus been replaced.
This is surprising as all maps are well known league play maps
or have been provided with the StarCraft Brood War distribution
itself.
Setup, replays and results for the CIG 2012 competition can be found
here\footnote{\url{http://ls11-www.cs.tu-dortmund.de/rts-competition/StarCraft-cig2012}}.

\begin{table}[t]
\caption{Results of the CIG 2012 competition.}
\label{tab:cig2012}
\centering
\begin{tabular}{|c|l|r|}
\hline
{\bfseries Position} & {\bfseries Bot} & {\bfseries Win \%} \\
\hline
1 & Skynet & 78.3\% \\
2 & UAlbertaBot & 65.2\% \\
3 & AIUR & 60.4\% \\
4 & Adjutant & 58.6\% \\
5 & Nova & 52.4\% \\ 
\hline
\end{tabular}
\end{table}

\begin{figure*}[tb]
    \centering
    \includegraphics[width=\textwidth]{figures/botLadder1.pdf}
    \caption{Bot's Elo Rating from February 1, 2012 to June 20, 2012}
    \label{fig:botLadder1}
\end{figure*}

\begin{figure*}[tb]
    \centering
    \includegraphics[width=\textwidth]{figures/botLadder2.pdf}
    \caption{Bot's Elo Rating from September 1, 2012 to January 10, 2013}
    \label{fig:botLadder2}
\end{figure*}

The overall results are displayed in table~\ref{tab:cig2012},
and the win rate evolution over time in figure~\ref{fig:cig2012-results}.
These are quite consistent with the results of the AIIDE 2012 competition,
so that we can conclude that the best bots are not very dependent on 
knowing the maps beforehand. However, the bot vs. bot win rates as
displayed in figure~\ref{fig:cig2012-mapresults} show some interesting
trends. On the maps with more possible start points, some bots do 
better than others, namely SCAIL, Adjutant, Nova, and UAlbertaBot,
the latter probably due to its very efficient scouting routine.
Some bots however suffer from the increased uncertainty about
the enemies' position, namely Xelnaga and BroodwarBotQ.
